---
title: "Do you trust your gut?"
author: "Jhonatan da Silva"
date: "September, 2018"
output: md_document
---

### Do you trust your gut? A probabilistic approach. 

I was having some conversations about the Monty Hall experiment and got quite surprised
on how many people even after the explanation, still think that they would have more chances of winning by sticking with the first choice. I started to think on ways that I could better show that this wasn't the fact indeed. Elaborating on this let me to this analysis that I present to you today.

First of all, let me give you an introducton about what this problem is all about, Monty Hall was a game show host, in this problem at hand we analyse the part "Let's make a deal" where contestants were asked to pick one out of three doors. Behind one of the doors, it was a prize, and behind the others a goat. The game worked in this way, it was asked to the contestant which of the doors did they choose. After the fact, it was showed that behind one of the doors there was a goat. Now we have only two doors, the one that had been picked, and the other that wasn't open. In this part of the game he ask the question. Do you want to change doors? Based on that, is in the best interest of the contestant switch doors? 

Let's look at the first facts, in the inicial phase, you have 3 options and just one prize to catch, so you have 1/3 chances of winning. You can prove mathematically that if the contestant switches the door after the fact, he now have 2/3 chances of winning.

### How could we go about proving this
I didn't follow the math path as my intention was giving some intuition on the problem at hand. The best way I could show that switching is in fact in the best interest, was to stop with all those peoples that I've talked and make this experiment. Stand by 3 doors, put a prize behind one of them, nothing in the others and following the game. I could do this 20 times, the first 10 times we wouldn't change doors, the next 10 we would. In that way I could show that it's really in the best interest switching doors.
But I bet you're also thinking what I did, there's no way in hell that a single person would've given me the time and pacience to make this experiment. Also if we just had made  20 times this game we wouldn't be so sure of the final results. So what's the other option? Simulations!

I ran a simulation for 10000 games and those are the results. Just what it was mathematically expected. If you stick by your door, you have about 1/3 chance of winning, if you've switched, it was about 2/3.

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(ggthemes)
```

```{r, echo=FALSE}

simulation_attempts <- 10000
monty_hall <- function(attempts,I_switch=FALSE){
  replicate(attempts,{ 
    doors <- as.character(seq(1,3))
    prize <- sample(c("house","goat","goat"))
    door_with_prize <- doors[prize %in% "house"]
    i_choose <- sample(doors,1)
    my_and_prize_door <- doors %in% c(i_choose,door_with_prize)
    what_door_to_show <- sample(doors[!my_and_prize_door],1)
    choice_that_have_left <- !doors%in%c(i_choose,what_door_to_show)
    if(I_switch) {
      final_say <- doors[choice_that_have_left]
    }
    else final_say <- i_choose
    final_say
    final_say == door_with_prize
  }) 
}
trusted_my_gut <- mean(monty_hall(simulation_attempts))*100
switched <- mean(monty_hall(simulation_attempts,I_switch=TRUE))*100

monty_hall_experiment=data.frame(name=c("Trusted my gut","Switched"),
                                 probability=c(trusted_my_gut,switched))
monty_hall_experiment %>% ggplot(aes(x=name, y=probability,label=probability,fill=probability)) + 
  geom_bar(stat = "identity") + geom_text(vjust=2,color='white') + 
  ylim(0,100) + ylab('Probability of winning') + xlab('') + 
  ggtitle('Monty Hall Experiment') +theme_economist()
```

### I Still trust my gut
For the ones that still are not true believers of the wonders of probability, I'll still try to convice you. I created another experiment called 10 cards experiment, it's pretty like the door problem. I have 10 cards, numered from 1 up to 10. I'll give you 10 bucks if you can choose the number 8 card. In the first place, you take a card and don't look at it, then I'll turn 8 cards up to you, none of these are the 8 card. Now the game is on, do you, or don't you switch your card? In the first choice you had 1/10 chances of winning, if you switch in the last move, you'll have 9/10 chances of winning! And here's the simulation 
```{r, echo=FALSE}
the_10_cards_experiment <- function(attempts,I_switch=FALSE){
  replicate(attempts,{ 
    cards <- seq(1,10)
    prize <- sample(cards,1)
    i_choose <- sample(cards,1)
    my_card_and_prize <- cards %in% c(i_choose,prize)
    what_cards_to_show <- sample(cards[!my_card_and_prize],8)
    choice_that_have_left <- !cards%in%c(i_choose,what_cards_to_show)
    if(I_switch) {
      final_say <- cards[choice_that_have_left]
    }
    else final_say <- i_choose
    final_say == prize
  }) 
}
trusted_my_gut <- mean(the_10_cards_experiment(simulation_attempts))*100
switched <- mean(the_10_cards_experiment(simulation_attempts,I_switch=TRUE))*100

monty_hall_experiment=data.frame(name=c("Trusted my gut","Switched"),
                                 probability=c(trusted_my_gut,switched))
monty_hall_experiment %>% ggplot(aes(x=name, y=probability,label=probability,fill=probability)) + 
  geom_bar(stat = "identity") + geom_text(vjust=1.3,color='white') + 
  ylim(0,100) + ylab('Probability of winning') + xlab('') + 
  ggtitle('The 10 Cards Experiment') +theme_economist()
```

### Exaggerating the problem
Now let's suppose I couldn't convice you so far, and suppose you are still reading up to this point. Let's go crazy and exaggerate the problem. Now I have a deck of 20 cards, or 30 cards, 50? And what about 70? 100? I created a simulation from 3 cards up to 100 and here are the results, About 50 cards you pretty much win every time if you switches your card and lose if you don't! Crazy? No, just probability. Hope you've liked it 
```{r,echo=FALSE}
the_n_cards_experiment <- function(attempts,number_of_cards,I_switch=FALSE){
  replicate(attempts,{ 
    cards <- seq(1,number_of_cards)
    prize <- sample(cards,1)
    i_choose <- sample(cards,1)
    my_card_and_prize <- cards %in% c(i_choose,prize)
    what_cards_to_show <- cards[!my_card_and_prize]
    if(length(what_cards_to_show) > 1){
      what_cards_to_show <- sample(what_cards_to_show,number_of_cards-2)
    }
    choice_that_have_left <- !cards%in%c(i_choose,what_cards_to_show)
    if(I_switch) {
      final_say <- cards[choice_that_have_left]
    } 
    else final_say <- i_choose
    final_say == prize
  }) 
}
number_of_cards <- seq(3,100)
trusted_my_gut <- c()
switched <- c()
aux <- 1
for(counter in number_of_cards){
  trusted_my_gut[aux] <- mean(the_n_cards_experiment(10000,counter))*100
  aux <- aux + 1
}
aux <- 1
for(counter in number_of_cards){
  switched[aux] <- mean(the_n_cards_experiment(10000,counter,I_switch=TRUE))*100
  aux <- aux + 1
}

n_cards_experiment=data.frame(number=number_of_cards,trusted=trusted_my_gut,
                              switched=switched)
n_cards_experiment %>% ggplot(aes(x=n_cards_experiment$number)) +
  geom_point(y=n_cards_experiment$switched,alpha=.5,stat='identity') +
  ylim(0,100) + ylab('aa of winning') + xlab('a') + 
  ggtitle('The N Cards Experiment') + theme_fivethirtyeight()
```

### For the curious enogh
Here's the related code
[Do you trust your gut?](https://github.com/silvajhonatan/do-you-trust-your-gut)